Supplementary Information -- Morphological convergence as on-line lexical analogy

### 2020-01-29 supplementary_information/si.Rnw
some progress

# libs
library(tidyverse)
library(xtable)
library(lme4)

@@

@@

splive_gcm = test2 %>% 
  mutate(
    r_baseline_gcm_features = scales::rescale(baseline_gcm_features)
  ) %>% 
  filter(word == 'splive') %>% 
  pull(r_baseline_gcm_features) %>% 
  unique %>% 
  round(2)

@@

# get the raw output of the baseline feature-distance mgl
mgl_baseline_rules_raw = read_tsv('~/Github/RaczBecknerHayPierrehumbert2019/models/mgl/baseline_mgl/CELEXFull3.sum') %>% 
  mutate(
    form1 = str_replace_all(form1, '»', ''), # get rid of syllable boundary markers
    form2 = str_replace_all(form2, '»', '')
  )

@@

verb_handles = read_csv('~/Github/RaczBecknerHayPierrehumbert2019/data/256verbs_print_forms.txt')

@@

<<main_model1, echo=F,eval=T>>=
test2 = test2 %>% 
  mutate(
    verb_baseline_average = baseline_av - mean(baseline_av, na.rm = T),
    participant_prereg_average = pre_reg_av - mean(pre_reg_av, na.rm = T),
    bot_lexical_typicality = case_when(
        lex_typicality == 'typical' ~ 'typical',
        lex_typicality == 'random' ~ 'random',
        lex_typicality == 'atypical' ~ 'reversed'
      ) %>% 
      factor(levels = c('typical', 'random', 'reversed')),
    bot_regularisation_rate = reg_rate %>% 
      factor(levels = c('nc', '-40%', '+40%'))
  )

@@

@@

save(lme.fit1, file = 'si_fit1.rda')
vif.mer(lme.fit1)
@

@@

# summary(fit_power1)
# summary(fit_power2)

@@

baseline = baseline %>% 
  mutate(
    r_baseline_mgl_features = scales::rescale(baseline_mgl_features),
    r_baseline_gcm_features = scales::rescale(baseline_gcm_features),
  )

@@

test2 = test2 %>% 
 mutate(
    r_baseline_mgl_features = scales::rescale(baseline_mgl_features),
    r_baseline_gcm_features = scales::rescale(baseline_gcm_features),
    r_individual_mgl_features = scales::rescale(individual_mgl_features),
    r_individual_gcm_features = scales::rescale(individual_gcm_features)
 )

@@

\subsection{Extra information provided by the individual model over the \sc{Celex}-model}

Second, we considered the amount of extra variation explained by the individual-model over the \sc{Celex}-model, for the MGL and the GCM, respectively.

Section 6.4 in the main text explains that we calculate the extra information given by the individual models by subtracting the \sc{Celex} model prediction from the individual model prediction for each item for each participant.

@@

test2 = test2 %>% 
  mutate(
    gcm_score_diff = r_individual_gcm_features - r_baseline_gcm_features,
    mgl_score_diff = r_individual_mgl_features - r_baseline_mgl_features
)

resp_post_reg ~ 1 + verb_baseline_average * bot_lexical_typicality + participant_prereg_average + bot_regularisation_rate

indiv_m5 = glmer(resp_post_reg ~ r_baseline_gcm_features + gcm_score_diff + participant_prereg_average + 
                   (1 + r_baseline_gcm_features + gcm_score_diff | participant_id) + (1 | word), 
                 data = test2, 
                 family = binomial (link = logit), 
                 control=glmerControl(optimizer="bobyqa"))

@@

indiv_m6 = glmer(resp_post_reg ~ r_baseline_mgl_features + mgl_score_diff + participant_prereg_average  + 
                   (1 + r_baseline_mgl_features + mgl_score_diff | participant_id) + (1 | word), 
                 data = test2, 
                 family = binomial (link = logit), 
                 control=glmerControl(optimizer="bobyqa"))

@@

These techniques are indicative of the relative strengths of the individual-MGL and -GCM respective to their \sc{Celex}-counterparts and each other. However, they do not compare them directly using statistical tests.

@@

>>>>

# libs
library(tidyverse)
library(xtable)
library(lme4)
library(patchwork) # take this out

@@

# setting up columns in baseline and post-test data
baseline = baseline %>% 
  mutate(
    r_baseline_mgl_features = scales::rescale(baseline_mgl_features), # rescale regularity weights
    r_baseline_gcm_features = scales::rescale(baseline_gcm_features),
  )

test2 = test2 %>% 
  mutate(
    verb_baseline_average = baseline_av - mean(baseline_av, na.rm = T), # center verb average
    participant_prereg_average = pre_reg_av - mean(pre_reg_av, na.rm = T), # center participant average
    bot_lexical_typicality = case_when( # set factor levels used in paper, order factor
        lex_typicality == 'typical' ~ 'typical',
        lex_typicality == 'random' ~ 'random',
        lex_typicality == 'atypical' ~ 'reversed'
      ) %>% 
      factor(levels = c('typical', 'random', 'reversed')),
    bot_regularisation_rate = reg_rate %>% # order factor
      factor(levels = c('nc', '-40%', '+40%')),
    r_baseline_mgl_features = scales::rescale(baseline_mgl_features), # scale regular weights
    r_baseline_gcm_features = scales::rescale(baseline_gcm_features),
    r_individual_mgl_features = scales::rescale(individual_mgl_features),
    r_individual_gcm_features = scales::rescale(individual_gcm_features),
    gcm_score_diff = r_individual_gcm_features - r_baseline_gcm_features, # calculate score diff
    mgl_score_diff = r_individual_mgl_features - r_baseline_mgl_features
  )

# mgl illustrations

## get the raw output of the baseline feature-distance mgl
mgl_baseline_rules_raw = read_tsv('~/Github/RaczBecknerHayPierrehumbert2019/models/mgl/baseline_mgl/CELEXFull3.sum') %>% 
  mutate(
    form1 = str_replace_all(form1, '»', ''), # get rid of syllable boundary markers
    form2 = str_replace_all(form2, '»', '')
  )

## get verb handles
verb_handles = read_csv('~/Github/RaczBecknerHayPierrehumbert2019/data/256verbs_print_forms.txt')

@@

splive_gcm = test2 %>% 
  filter(word == 'splive') %>% 
  pull(r_baseline_gcm_features) %>% 
  unique %>% 
  round(2)

@@

@@

@@

<<main_model1, echo=F,eval=T>>=

@@

# the model reported in the paper

@@

# the first comparison that comes to mind (no interaction):

lme.fit1b = glmer(resp_post_reg ~ 
    1 + verb_baseline_average + bot_lexical_typicality + participant_prereg_average + bot_regularisation_rate +
    (1 + verb_baseline_average | participant_id) + 
    (1  | word), 
  data = test2, 
  family = binomial (link = logit),
  control=glmerControl(optimizer="bobyqa")
)

# the second comparison that comes to mind (three-way interaction):

lme.fit1c = glmer(resp_post_reg ~ 
    1 + verb_baseline_average * bot_lexical_typicality * bot_regularisation_rate + participant_prereg_average + 
    (1 + verb_baseline_average | participant_id) + 
    (1  | word), 
  data = test2, 
  family = binomial (link = logit),
  control=glmerControl(optimizer="bobyqa", optCtrl = list(maxfun = 100000))
)

save(lme.fit1, file = 'si_fit1.rda')
save(lme.fit1b, file = 'si_fit1b.rda')
save(lme.fit1c, file = 'si_fit1c.rda')
vif.mer(lme.fit1)
vif.mer(lme.fit1c)

anova(lme.fit1,lme.fit1b,lme.fit1c)

@@

# summary(fit_power1)
# summary(fit_power2)
save(fit_power1, file = 'si_fit_power1.rda')
save(fit_power2, file = 'si_fit_power2.rda')

@@

@@

@@

\subsection{Extra information provided by the individual model over the {\sc Celex}-model}

Second, we considered the amount of extra variation explained by the individual-model over the {\sc Celex}-model, for the MGL and the GCM, respectively.

Section 6.4 in the main text explains that we calculate the extra information given by the individual models by subtracting the {\sc Celex} model prediction from the individual model prediction for each item for each participant.

@@

indiv_m5 = glmer(resp_post_reg ~
    1 + r_baseline_gcm_features + gcm_score_diff + participant_prereg_average + 
    (1 + r_baseline_gcm_features + gcm_score_diff | participant_id) + (1 | word), 
                 data = test2, 
                 family = binomial (link = logit), 
                 control=glmerControl(optimizer="bobyqa"))

@@

indiv_m6 = glmer(resp_post_reg ~ 
    1 + r_baseline_mgl_features + mgl_score_diff + participant_prereg_average  + 
    (1 + r_baseline_mgl_features + mgl_score_diff | participant_id) + (1 | word), 
                 data = test2, 
                 family = binomial (link = logit), 
                 control=glmerControl(optimizer="bobyqa"))

@@

These techniques are indicative of the relative strengths of the individual-MGL and -GCM respective to their {\sc Celex}-counterparts and each other. However, they do not compare them directly.

@@

<<comparison_model1, echo=F,eval=F>>=

# the model reported in the paper 
comp_fit1 = glmer(resp_post_reg ~ 
    1 + r_baseline_mgl_features + r_individual_gcm_features + participant_prereg_average  + 
    (1 + r_baseline_mgl_features + r_individual_gcm_features | participant_id) + (1 | word), 
                 data = test2, 
                 family = binomial (link = logit), 
                 control=glmerControl(optimizer="bobyqa"))

# model without baseline mgl
comp_fit1b = glmer(resp_post_reg ~ 
    1 + r_individual_gcm_features + participant_prereg_average  + 
    (1 + r_individual_gcm_features | participant_id) + (1 | word), 
                 data = test2, 
                 family = binomial (link = logit), 
                 control=glmerControl(optimizer="bobyqa"))

# model without individual gcm
comp_fit1c = glmer(resp_post_reg ~ 
    1 + r_baseline_mgl_features + participant_prereg_average  + 
    (1 + r_baseline_mgl_features | participant_id) + (1 | word), 
                 data = test2, 
                 family = binomial (link = logit), 
                 control=glmerControl(optimizer="bobyqa"))

# model with individual mgl
comp_fit1d = glmer(resp_post_reg ~ 
    1 + r_baseline_mgl_features + r_individual_gcm_features + r_individual_mgl_features + participant_prereg_average  + 
    (1 + r_baseline_mgl_features + r_individual_gcm_features | participant_id) + (1 | word), 
                 data = test2, 
                 family = binomial (link = logit), 
                 control=glmerControl(optimizer="bobyqa"))

# model with celex gcm
comp_fit1e = glmer(resp_post_reg ~ 
    1 + r_baseline_mgl_features + r_individual_gcm_features + r_baseline_gcm_features + participant_prereg_average  + 
    (1 + r_baseline_mgl_features + r_individual_gcm_features | participant_id) + (1 | word), 
                 data = test2, 
                 family = binomial (link = logit), 
                 control=glmerControl(optimizer="bobyqa"))

save(comp_fit1, file = 'si_comp_fit1.rda')
save(comp_fit1b, file = 'si_comp_fit1b.rda')
save(comp_fit1c, file = 'si_comp_fit1c.rda')
save(comp_fit1d, file = 'si_comp_fit1d.rda')
save(comp_fit1e, file = 'si_comp_fit1e.rda')

We used the {\sc Celex}-GCM, {\sc Celex}-MGL, individual-GCM, and individual-MGL together to predict participant post-test responses. The best model, reported in the paper, contains the {\sc Celex}-MGL and the individal-GCM. The summary can be seen in Table \ref{comp3}.

<<comparison_model2, echo=F,eval=T>>=

load('si_comp_fit1.rda')
load('si_comp_fit1b.rda')
load('si_comp_fit1c.rda')
load('si_comp_fit1d.rda')
load('si_comp_fit1e.rda')

chisq1 = anova(comp_fit1, comp_fit1b)[2,6] %>% round(2) # without baseline mgl
chisq2 = anova(comp_fit1, comp_fit1c)[2,6] %>% round(2) # without individual gcm
chisq3 = anova(comp_fit1, comp_fit1d)[2,6] %>% round(2) # with individual mgl
chisq4 = anova(comp_fit1, comp_fit1e)[2,6] %>% round(2) # with celex gcm

<<comparison_model3, echo=F,eval=T, results=tex>>=
print(xtable(summary(comp_fit1)$coef, digits = 3, label = 'comp3', caption = 'Celex-MGL and Individual-GCM predict variation in the post-test'))
@

Goodness-of-fit tests show that model fit is significantly worse if we exclude {\sc Celex}-MGL ($\chi^2 = $ \Sexpr{chisq1}) or individual-GCM ($\chi^2 = $ \Sexpr{chisq2}) as predictors. It is not significantly better if we include individual-MGL ($\chi^2 = $ \Sexpr{chisq3}) or {\sc Celex}-GCM ($\chi^2 = $ \Sexpr{chisq4}).

\subsection{Visualising model contributions to explaining variation in post-test responses}

Figure \ref{bigfigure} shows how baseline and post-test regularisation means of words compare for participant responses, the GCM, and the MGL.

<<model_figure, echo=F,eval=T>>=

# can't figure first panel out

first_panel = test2 %>% 
  group_by(
    word,
    bot_lexical_typicality,
    baseline_av
  ) %>%
  summarise(
    post_reg_mean = post_reg_av %>% 
      mean()
  ) %>% 
  ungroup() %>% 
  mutate(
    baseline = scales::rescale(baseline_av),
    posttest = scales::rescale(post_reg_mean),
    type = 'Participant responses'
  )

second_panel = test2 %>% 
  group_by(
    word,
    bot_lexical_typicality,
    r_baseline_gcm_features
  ) %>% 
  rename(
    baseline = r_baseline_gcm_features
  ) %>% 
  summarise(
    posttest = mean(r_individual_gcm_features)
  ) %>% 
  mutate(
    type = 'GCM'
  )

third_panel = test2 %>% 
  group_by(
    word,
    bot_lexical_typicality,
    r_baseline_mgl_features
  ) %>% 
  rename(
    baseline = r_baseline_mgl_features
  ) %>% 
  summarise(
    posttest = mean(r_individual_mgl_features)
  ) %>% 
  mutate(
    type = 'MGL'
  )

# three_studies = bind_rows(first_panel,second_panel,third_panel)  
three_studies = bind_rows(second_panel,third_panel)

p1 = ggplot(three_studies, aes(x = baseline, y = posttest, colour = bot_lexical_typicality, shape = bot_lexical_typicality, linetype = bot_lexical_typicality)) +
  guides(size = F) +
  geom_point() +
  geom_smooth(se = F, method = 'loess', formula = 'y ~ x') +
  facet_wrap( ~ type) +
  scale_shape_discrete(name="Lexical\ntypicality") +
  # scale_colour_discrete(name="Lexical\ntypicality") +
  xlab('verb baseline regularization') +
  ylab('verb ESP regularization') +
  guides(linetype = F) +
  scale_colour_brewer(palette = 'Set2', name="Lexical\ntypicality")

\begin{figure}
<<model_figure2, fig=T,echo=F,eval=T, height = 4, width = 13>>=
wrap_elements(grid::textGrob("I'm a \n giant \n turnip")) + p1
@
\caption{Baseline and individual regularisation weights: participants, GCM, MGL}
\label{bigfigure}
\end{figure}

\newpage
